---
title: "0 - Details and Setup"
---

## Create folders for output

```{r}
#| warning: false
dir.create("Data/Intermediate")
dir.create("Data/Datasets")
dir.create("Data/Final")
dir.create("Data/Spatial")
dir.create("Results")
dir.create("Results/Figures")
```

## Get packages

If possible, use renv to restore the packages
```{r}
#| eval: false
renv::restore()
```

Otherwise install by hand e.g., `install.packages(c("tidyverse", "sf")`


## Establish date ranges

We want May 1st to July 31st in Julian days

```{r}
lubridate::yday(c("2023-05-01",
                  "2020-05-01", # Leap year
                  "2023-07-31",
                  "2020-07-31")) # Leap year
```

We'll use 121 - 213


## Acquire raw data files

- E-bird data must be requested [here](https://ebird.org/data/download/)
  - I made a special request for Long-billed Curlews from 2010-05 to 2022-08
  - Once the request is approved you'll get an email and can download the data
    (I've found this can take between 2 hours and 5 days)
- You'll also need to download the complete checklist data (currently not available as a special request)
- Extract all the files (including the internal compressed files)
  - You'll have to move the 2 main data files into `Data/Raw` after they've been extracted
  - The sampling file: `ebd_sampling_relMay-2023.txt`
  - The observations file: `ebd_lobcur_201005_202208_relMay-2023.txt`
- Tell `auk` where that folder is (below) and **restart R**
  - Only need to do this step once

```{r}
auk::auk_set_ebd_path(path = "Data/Raw/", overwrite = TRUE)
```


