---
title: "6 - Figures"
cache: true
---

This document contains grid maps of each year showing the observed curlew grids
against the non-observed grids. I've put on the max/min lat/lon and the centroids
for these observed grids as a sanity check to make sure it looks like the calculations are 
accurate.

These maps show quality controled data (i.e. >= 5 or 10 years with a checklist), 
but are not filtered by distance percentiles (because I think you shouldn't :P).

**To see an image more closely, right-click and select "Open Image in new Tab" (or similar)**


```{r}
library(tidyverse)
library(sf)
library(patchwork)

base_map <- read_rds("Data/Datasets/base_map.rds")
bcr <- read_rds("Data/Datasets/bcr_map.rds")
bc <- read_rds("Data/Datasets/bc_map.rds")

final <- read_rds("Data/Datasets/final.rds") |>
  filter(region == "all", grid_perc == "grids_100", trt == "all") |>
  select(area_ha, year, grids)

present_measures <- read_rds("Data/Datasets/final.rds") |>
  filter(region == "all", grid_perc == "grids_100", trt == "present") |>
  select(area_ha, year, min_lon, min_lat, max_lon, max_lat, cent_lon, cent_lat)

plot <- left_join(final, present_measures)

```

## 10x10 grids by year
- Box defines the min/max lat/lon for the grids with *observed* curlews
- The dark squiggle at the top is British Columbia (only the BCR regions of interest)
- The purple dots are the centroids of the grids with *observed* curlews

```{r}
#| fig-dpi: 300
#| fig-width: 12
#| fig-asp: 0.8
g <- plot |>
  filter(area_ha == 10000) |>
  pmap(~ggplot(..3, ) +
         geom_sf(data = bcr, fill = "white") +
         geom_sf(data = bc, fill = NA, linewidth = 1) +
         geom_sf(aes(fill = presence, colour = presence)) + 
         coord_sf(crs = 4326) +
         geom_hline(yintercept = c(..5, ..7)) +
         geom_vline(xintercept = c(..4, ..6)) +
         geom_point(x = ..8, y = ..9, aes(colour = "Centroid"), size = 4) +
         scale_fill_manual(values = c("Centroid" = "#440154",
                                      "FALSE" = "#FDE725",
                                      "TRUE" = "#21908C"), aesthetics = c("colour", "fill")) +
         labs(title = ..2, fill = "Curlews Present", colour = "Curlews Present",
              caption = "10x10km grid; BC outlined in bold")
  )

walk(g, print)
```

## 20x20 grids by year
- Box defines the min/max lat/lon for the grids with *observed* curlews
- The dark squiggle at the top is British Columbia (only the BCR regions of interest)
- The purple dots are the centroids of the grids with *observed* curlews
```{r}
#| fig-dpi: 300
#| fig-width: 12
#| fig-asp: 0.8
g <- plot |>
  filter(area_ha == 40000) |>
  pmap(~ggplot(..3, ) +
         geom_sf(data = bcr, fill = "white") +
         geom_sf(data = bc, fill = NA, linewidth = 1) +
         geom_sf(aes(fill = presence, colour = presence)) + 
         coord_sf(crs = 4326) +
         geom_hline(yintercept = c(..5, ..7)) +
         geom_vline(xintercept = c(..4, ..6)) +
         geom_point(x = ..8, y = ..9, aes(colour = "Centroid"), size = 4) +
         scale_fill_manual(values = c("Centroid" = "#440154",
                                      "FALSE" = "#FDE725",
                                      "TRUE" = "#21908C"), aesthetics = c("colour", "fill")) +
         labs(title = ..2, fill = "Curlews Present", colour = "Curlews Present",
              caption = "20x20km grid; BC outlined in bold")
  )

walk(g, print)
```
